---
title: "Assignment 2: Coding Basics"
author: Rachel Schoenecker
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
  # Creating sequence from 1 to 100 by 4's and naming it "seq.4"
  seq.4 <- seq(1, 100, 4) 
  seq.4

#2. 
  # Calculating mean and median of seq.4
  mean(seq.4); median(seq.4) 

#3. 
  # Logical statement asking whether the mean of seq.4 is greater than the median
  mean(seq.4) > median(seq.4) 
```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5a.
  stud.names <- c("Blair", "Serena", "Dan", "Nate") # stud.names is a character vector
  stud.names
#5b.
  stud.grades <- c(98, 74, 91, 48) # stud.grades is a numeric vector
  stud.grades
#5c.
  stud.pass <- c(T, T, T, F) # stud.pass is a logical vector
  stud.pass
#7.
  gradebook <- data.frame(stud.names, stud.grades, stud.pass)
  gradebook
#8.
  colnames(gradebook) <- c("Name", "Grade", "Pass")
  gradebook
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: Both are two-dimensional features containing data, but a data frame can include data of many different types (e.g., numeric, character, and logical) while a matrix can only contain one type of data (e.g., numeric or character).

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10.
pass.test <- function(grade){
  outcome <- ifelse(grade >= 50, T, F)
  print(outcome)
}

#11.
  pass.test(gradebook$Grade)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The 'ifelse' option worked because it can evaluate multiple values in a vector in one step while the if/else statement can only evaluate one value at a time. In order to use if/else for a vector of multiple values a 'for' loop would need to be employed as well.
